---
title: "System Dynamics Simulation in WimpTools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{System Dynamics Simulation in WimpTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Beyond mapping static states, `WimpTools` incorporates an engine for mathematically simulating systemic change. Because the formulation of a wimp grid requires answers to the question *"If X changed, what would happen to Y?"* (Sanfeliciano et al., 2024), the `scenariomatrix()` engine models these responses over iterative time-steps, exposing long-term homeostatic resistance, equilibrium paths, or system cascades.

### References

*   Sanfeliciano, A., et al. (2024). *The Weighted Implications Grid: A Graph-Theoretical Approach to Modelling Psychological Change Construction*.

For all examples below, we will utilize the `example_wimp` package dataset.
```{r setup}
library(WimpTools)
data(example_wimp)
```

## Simulating Psychological Dynamics (PCSD)

### `scenariomatrix()`

**Purpose:** 
The computational core of the simulation. It repeatedly applies the weighted implication matrix against the psychological system state to calculate downstream cognitive cascading. 

**Usage and Arguments:**
```{r scene}
# Defining an initial psychological activation ("pushing" a construct)
# Must be exactly length N. Zeros imply no external activation.
external_change <- rep(0, example_wimp$global$n_constructs) 
external_change[1] <- 0.5 # Activating the first construct

example_scenario <- scenariomatrix(wimp = example_wimp, 
                              act_vector = external_change, 
                              infer = "self dynamics",
                              max_iter = 10)
```

* **`wimp`**: A valid `wimp` object.
* **`act_vector`**: A numeric vector exactly the length of $N$ constructs. This sets the initial "external" activation applied to the system at $t=1$.
* **`infer`**: Mathematical methodology defining exactly *what* is simulated. 
  * `"self dynamics"`: (Default). The standard algorithm. Computes how the *Actual Self* will shift and settle across iterations under an activating stressor.
  * `"impact dynamics"`: An alternative approach mapping raw cascade impact dispersion.
* **`thr`**: The decay thresholding rule used to simulate real-world psychological friction. Options: `"saturation"` (default limits growth logarithmically bounding output space), `"tanh"`, or `"none"` (linear endless propagation). 
* **`max_iter`**: Limit of iterations in the simulation $t_1 \rightarrow t_{max}$.
* **`e`**: Convergence tolerance factor. If system changes stay below `e` for `stop_iter` periods, simulation halts early.
* **`exclude_dilemmatics`**: (`logical`) If true, isolates dilemmatic constructs out of the equilibrium calculations.

**Output:** 
Returns an S3 object of class `scn`. It contains a nested matrix where columns correspond to the constructs and rows correspond to simulation iterations `iter 0:max_iter`.

## Analyzing Simulation Data

Once `scenariomatrix()` returns an `scn` object, several tools extract interpretation metrics.

### `pcsd()`
**Purpose:** Personal Constructs System Dynamics plot. Takes the numeric `scn` matrix and builds an interactive Plotly line graph detailing the longitudinal shift in magnitude (Y-axis) of each construct against the unfolding iterative time-steps (X-axis). 

```{r pcsd, fig.width=7, fig.height=6}
pcsd(scn = example_scenario, vline = NA)
```

### `auc_index()`
**Purpose:** Computes the standardized geometric Area Under the Curve (AUC) for each construct's temporal line graph from the PCSD simulation. Provides an aggregate index of long-term activation.
```{r auc}
auc_index(scn = example_scenario, std = TRUE)
```

### `stability_index()`
**Purpose:** Calculates the standard deviation vector for each construct. Constructs possessing exceptionally high standard deviation scores exhibited massive oscillations during the simulation.
```{r stab}
stability_index(scn = example_scenario)
```

### `pcsd_summary()`
**Purpose:** Convenient helper function returning a consolidated matrix containing: `Initial` state (t=0), `Final` state (t=max), `Average` activation level, and total `Difference` generated by the simulation.
```{r pcsds}
pcsd_summary(scn = example_scenario)
```
