---
title: "Psychological Well-Being and Adjustment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bienestar Psicologico: Psychological Well-Being and Adjustment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In the clinical application of Personal Construct Psychology, psychological well-being is strongly associated with the alignment between how a person perceives themselves (Actual Self) and how they wish to be (Ideal Self). `WimpTools` offers specifically tailored indices to calculate and visualize this alignment, accounting for the complex topological nature of the Weighted Implication Grid (Sanfeliciano et al., 2024).

This vignette details the functions used to extract indices of psychological adjustment, map out congruence profiles, and monitor therapeutic progress across time. In particular, the core self-adjustment metrics are based on the Similarity Self/Ideal Index (SSI) (Sanfeliciano et al., 2025).

### References

*   Sanfeliciano, A., et al. (2024). *The Weighted Implications Grid: A Graph-Theoretical Approach to Modelling Psychological Change Construction*.
*   Sanfeliciano, A., Hurtado-Martínez, C., Botella García del Cid, L., & Saúl, L. A. (2025). Similarity Self/Ideal Index (SSI): A Feature-Based Approach to Modeling Psychological Well-Being. *Mathematics*.

## Adjustment Indices

### `construct_index()`

**Purpose:** 
Calculates the systemic proportion of construct types based solely on the mathematical relationship between the self and ideal poles. 
A clinically well-adjusted individual generally exhibits a high proportion of "Congruent" constructs and a low proportion of "Discrepant" or "Dilemmatic" constructs.

**Usage and Arguments:**
```{r construct_idx}
library(WimpTools)
construct_index(wimp = example_wimp)
```
* **`wimp`**: A valid `wimp` S3 object.

**Output:** 
A matrix with two columns: `Frequency` (absolute count) and `Proportion` (percentage of total $N$). It rows correspond to the four core states: `Congruents`, `Discrepants`, `Dilemmatics`, and `Undefined`.

### `self_index()`

**Purpose:** 
Computes the mathematical similarity between the Actual Self, the Ideal Self, and all generated Hypothetical Scenarios. The core output is the **Self/Ideal correlation**, which serves as the primary numeric proxy for psychological adjustment.

By default, this calculates the **Semantic Similarity Index (SSI)**, a sophisticated algorithm designed specifically for WimpGrids that incorporates two cognitive biases: the salience of self-discrepancy (how much the person notices their flaws) and the salience of ideal aspiration (how much the person notices their desired goals).

**Usage and Arguments:**
```{r self_idx}
self_data <- self_index(wimp = example_wimp, method = "ssi", 
                        rc = TRUE, alpha = 0.5, beta = 0.5)
```
* **`wimp`**: A valid `wimp` object.
* **`method`**: Defaults to `"ssi"`. It can be swapped to `"pearson"`, `"kendall"`, or `"spearman"` if traditional correlational metrics are preferred.
* **`rc`**: (`logical`) If `TRUE` (default), utilizing Cohen's $r_c$, a reflection-invariant correlation measure crucial in repertory grid analysis because left/right pole orientation is often arbitrary.
* **`alpha`**: (`numeric`) Discrepancy salience parameter for SSI. Bounded $[0, 1]$. Represents attention allocated to the Actual Self. Default is $0.5$.
* **`beta`**: (`numeric`) Aspiration salience parameter for SSI. Bounded $[0, 1]$. Represents attention allocated to the Ideal Self. Default is $0.5$.

**Output:** 
Returns a `self_index` S3 list containing:
1. `global`: A data frame with three systemic similarities: `Self/Ideal` (Clinical Adjustment), `Self/Hypo`, and `Ideal/Hypo`.
2. `construct`: A data frame listing every generated hypothetical scenario alongside its structural `Congruence Scenario`, `SHS` (Self-Hypothetical Similarity), and `SHI` (Self-Ideal Similarity).
3. `structural_coefs`: Output of the SSI structural algorithms (if `method="ssi"`).

## Visualizing the Self

### `self_plot()`

**Purpose:** 
Produces an interactive radar chart overlaying the individual's "Actual Self" and "Ideal Self" across all constructs in the grid. It offers immediate, intuitive insights into where the major discrepancies exist.

**Usage:**
```{r self_plt, fig.width=7, fig.height=6}
self_plot(wimp = example_wimp)
```

**Output Details:** 
An interactive Plotly widget. 
* The **Green Dashed Line** represents the Ideal Self stringently mapped across poles.
* The **Purple Polygon** represents the Actual Self.
* Nodes are colored dynamically: Green = Congruent; Red = Discrepant; Yellow = Dilemmatic.

### `hypo_plot()`

**Purpose:** 
An interactive scatter plot mapping all psychological constructs onto a 2D Cartesian plane defined by their hypothetical similitudes. The X-axis represents the Self-Hypothetical Similarity (SHS) and the Y-axis represents the Self-Ideal Similarity (SHI).

**Usage and Arguments:**
```{r hypo_plt, fig.width=7, fig.height=6}
hypo_plot(wimp = example_wimp, text_size = 1.2, show_labels = TRUE)
```
* **`text_size`**: Determines the font sizes of text annotations.
* **`show_labels`**: (`logical`) If `TRUE`, displays construct names using a smart anti-collision algorithm. You can turn this to `FALSE` for highly cluttered grids to rely instead on hover-tooltips.
* **`...`**: Any extra arguments are passed natively to `self_index()` (e.g., `alpha`, `beta`).

### `ssi_heatmap()`

**Purpose:** 
Visualizes how the individual's psychological adjustment (Self/Ideal SSI score) varies across the entire theoretical parameter space of $\alpha$ (Discrepancy Salience) and $\beta$ (Aspiration Salience).

**Usage and Arguments:**
```{r ssi_heat, fig.width=7, fig.height=6}
ssi_heatmap(wimp = example_wimp, estimation = TRUE, palette = "viridis")
```
* **`estimation`**: (`logical`) If `TRUE`, the function calculates the structural preference weights to overlay a 2D probability distribution contour on the heatmap, identifying the most statistically probable psychological posture of the individual.
* **`palette`**: The color progression applied to the heatmap.

## Monitoring Therapeutic Progress

If multiple WimpGrids have been extracted from an individual across time (e.g., pre-intervention and post-intervention), `WimpTools` provides three robust functions to visualize the psychological change.

*Note: All monitoring functions automatically identify matching constructs between `wimp_t0` and `wimp_t1`. New constructs added in `t1` or constructs deleted from `t0` are omitted from the comparison to maintain mathematical integrity.*

### `monitoring_self()`
Plots two overlapping radar charts mimicking `self_plot()`. It maps both the Baseline Self and Actual Self, and Baseline Ideal and Actual Ideal, giving visual feedback on longitudinal directional movements.

```{r mon_self, fig.width=7, fig.height=6}
monitoring_self(wimp_t0 = example_wimp, wimp_t1 = su_wimp)
```

### `monitoring_ssi()`
Generates a side-by-side subplot of `ssi_heatmap()` for the pre-intervention and post-intervention evaluations.

```{r mon_ssi, fig.width=10, fig.height=5}
monitoring_ssi(wimp_t0 = example_wimp, wimp_t1 = su_wimp, estimation = TRUE)
```

### `monitoring_pb()`
Plots the longitudinal displacement of each psychological construct within the Presence-Balance index space. Constructs from `t0` are represented as faded gray nodes connected by a trajectory line to their new, brightly colored positions in `t1`. It also calculates and plots the overall centroid movement.

```{r mon_pb, fig.width=7, fig.height=6}
monitoring_pb(wimp_t0 = example_wimp, wimp_t1 = su_wimp, show_centroid = TRUE)
```
